---
title: "Problem set 1"
author: "Enrico Manlapig"
output:
  html_document:
    theme: cosmo

---

```{r,echo = FALSE}
include_solutions <- FALSE
```

```{r, echo=include_solutions, eval = include_solutions}
long_object <- seq(0, 200, by = 2) 

letters <- sample(LETTERS, 100, replace = TRUE) # vector

who_likes_animals <- 
  data.frame(size = sample(c("tiny", "regular", "big", "enormous"),
                           prob = c(0.245, 0.5, 0.25, 0.005),
                           size = 300, replace = TRUE),
             date_of_birth = Sys.Date() - runif(300) * 3000,
             species = as.factor(sample(c("cat",
                                        "dog",
                                        "fish",
                                        "snake",
                                        "bunny"),
                                 prob = c(0.4, 0.4, 0.1, 0.05, 0.05),
                                 size = 300, replace = TRUE)))

what_is_this <- list(long_object, 
                     letters,
                     who_likes_animals)

save(long_object, letters, who_likes_animals, what_is_this,
     file = "../data/Topic01.RData")
```


Download the dataset `Topic01.RData` and put it in the root directory of your Load the dataset `Topic01.RData` by including the following chunk at the beginning of your RMarkdown file.  
```{r, echo=include_solutions, eval = include_solutions}
load("../datasets/Topic01.RData")
```




### Problem 1

1. Use R to determine the types of the following objects

    a. `who_likes_animals` 
    a. `what_is_this[[2]]`
    a. `what_is_this[[3]][3]`
    a. `letters`
    a. `long_object`
    
```{r echo=include_solutions, eval = include_solutions}
class(who_likes_animals)

class(what_is_this[[2]])

class(what_is_this[[3]][3])

class(letters)

class(long_object)
```


2. What are their dimensions?
```{r, echo=include_solutions, eval = include_solutions}
dim(who_likes_animals)

dim(what_is_this[[2]])

dim(what_is_this[[3]][3])

dim(letters)

dim(long_object)
```


3. What is the value in the 2nd row and 3rd column of `who_likes_animals`?
```{r, echo=include_solutions, eval = include_solutions}
who_likes_animals[2,3]
```


4. Show the first 15 rows of `who_likes_animals`

```{r, echo=include_solutions, eval = include_solutions}
head(who_likes_animals, 15)
```


5. Show the last 10 rows of `who_likes_animals`

```{r, echo=include_solutions, eval = include_solutions}
tail(who_likes_animals, 10)
```



###  Problem 2

Use R Markdown to describe your weekly schedule.   Use heading sections,  bullets, and numbered lists to describe your various activities.  Use different fonts to refer to specific class, building, and instructor names.


### Problem 3

Suppose you are want to make meal with your friends.  Write a short paragraph indicating what you would make and who you would invite.  Use bullets and enumeration to create the following:

1. A (character) vector of ingredients you want in your meal (ingredients)
1. A (numeric) vector of how many of each item you need (units)
1. A (numeric) vector of estimated unit prices of each item (price)
1. A data frame that combines the three vectors together (data)
1. A new column equal to the product of units and price columns (cost)
1. Sum this new column to estimate the total cost of the meal

### Problem 4

1. Knit your work into an HTML and Word document
1. Upload your knitted documents to Canvas